---
title: "Analysis of the Adverse Health and Economic Impacts of US Storms"
author: "Ujas Italia"
date: "05/06/2020"
output: html_document
---

## 1: Synopsis
The goal of the assignment is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

The following analysis investigates which types of severe weather events are most harmful on:

1. Health (injuries and fatalities) 
2. Property and crops (economic consequences)

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

## 2: Data Processing

### 2.1 Data Loading
```{r cache=TRUE, eval=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile = "repdata_data_StormData.csv.bz2")
data <- read.csv("repdata_data_StormData.csv.bz2")
```

```{r include=FALSE, cache=TRUE}
library(dplyr)
library(ggplot2)
library(reshape2)
data <- read.csv("repdata_data_StormData.csv.bz2")
```

### 2.2 Data looks like 
```{r cache=TRUE}
head(data, 3)
```

### 2.3: Data Subsetting
```{r cache=TRUE, results='hide'}
# selecting required coulmns only
subData <- subset(data, select = c(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))

# deleting the zero value rows
subDataHealth <- subData[subData$FATALITIES !=0 |
                           subData$INJURIES !=0, ]

subDataProp <- subData[subData$PROPDMG !=0 |
                           subData$CROPDMG !=0, ]
```


### 2.4 Data conversion 
```{r cache=TRUE, results='hide'}
#creating data for first problem
subDataHealth <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, data = subData, FUN=sum)
subDataHealth$PEOPLE_LOSS <- subDataHealth$FATALITIES + subDataHealth$INJURIES
subDataHealth <- subDataHealth[order(subDataHealth$PEOPLE_LOSS, decreasing = TRUE), ]
subDataHealth <- subDataHealth[1:10,]
bad_stuff <- melt(subDataHealth, id.vars="EVTYPE", variable.name = "bad_thing")

#converting PROPDMGEXP and CROPDMGEXP values
subData$PROPDMGEXP <- gsub("[Hh]", "2", subData$PROPDMGEXP)
 subData$PROPDMGEXP <- gsub("[Kk]", "3", subData$PROPDMGEXP)
 subData$PROPDMGEXP <- gsub("[Mm]", "6", subData$PROPDMGEXP)
 subData$PROPDMGEXP <- gsub("[Bb]", "9", subData$PROPDMGEXP)
 subData$PROPDMGEXP <- gsub("\\+", "1", subData$PROPDMGEXP)
 subData$PROPDMGEXP <- gsub("\\?|\\-|\\ ", "0",  subData$PROPDMGEXP)
 subData$PROPDMGEXP <- as.numeric(subData$PROPDMGEXP)

 subData$CROPDMGEXP <- gsub("[Hh]", "2", subData$CROPDMGEXP)
 subData$CROPDMGEXP <- gsub("[Kk]", "3", subData$CROPDMGEXP)
 subData$CROPDMGEXP <- gsub("[Mm]", "6", subData$CROPDMGEXP)
 subData$CROPDMGEXP <- gsub("[Bb]", "9", subData$CROPDMGEXP)
 subData$CROPDMGEXP <- gsub("\\+", "1", subData$CROPDMGEXP)
 subData$CROPDMGEXP <- gsub("\\-|\\?|\\ ", "0", subData$CROPDMGEXP)
 subData$CROPDMGEXP <- as.numeric(subData$CROPDMGEXP)

 #changing null values to 0
 subData$PROPDMGEXP[is.na(subData$PROPDMGEXP)] <- 0
 subData$CROPDMGEXP[is.na(subData$CROPDMGEXP)] <- 0

# adding variables by calculating total value of PROPDMGEXP and CROPDMGEXP
 subData <- mutate(subData, PROPDMGTOTAL = PROPDMG * (10 ^ PROPDMGEXP), CROPDMGTOTAL = CROPDMG * (10 ^ CROPDMGEXP))
 
# creating data for scond problem
Economic_data <- aggregate(cbind(PROPDMGTOTAL, CROPDMGTOTAL) ~ EVTYPE, data = subData, FUN=sum)
Economic_data$ECONOMIC_LOSS <- Economic_data$PROPDMGTOTAL + Economic_data$CROPDMGTOTAL
Economic_data <- Economic_data[order(Economic_data$ECONOMIC_LOSS, decreasing = TRUE), ]
Economic_data <- Economic_data[1:10,]
econ_consequences <- melt(Economic_data, id.vars="EVTYPE", variable.name = "Damage_Type")
```


## 3: Results

### 3.1: Events that are Most Harmful to Population Health
```{r cache=TRUE}
# Create chart
 healthChart <- ggplot(bad_stuff, aes(x=reorder(EVTYPE, value), y=value))

 # Plot data as bar chart
 healthChart = healthChart + geom_bar(stat="identity", aes(fill=bad_thing), position="dodge")

 # Format y-axis scale and set y-axis label
 healthChart = healthChart + ylab("Frequency Count")

 # Set x-axis label
 healthChart = healthChart + xlab("Event Type")

 # Rotate x-axis tick labels
 healthChart = healthChart + theme(axis.text.x = element_text(angle=45, hjust=1))

 # Set chart title and center it
 healthChart = healthChart + ggtitle("Top 10 US Killers") + theme(plot.title = element_text(hjust = 0.5))

 print(healthChart)
```

### 3.2: Events that have the Greatest Economic Consequences
```{r cache=TRUE}
 econChart <- ggplot(econ_consequences, aes(x=reorder(EVTYPE, -value), y=value))

 # Plot data as bar chart
 econChart = econChart + geom_bar(stat="identity", aes(fill=Damage_Type), position="dodge")

 # Format y-axis scale and set y-axis label
 econChart = econChart + ylab("Cost (dollars)")

 # Set x-axis label
 econChart = econChart + xlab("Event Type")

 # Rotate x-axis tick labels
 econChart = econChart + theme(axis.text.x = element_text(angle=45, hjust=1))

 # Set chart title and center it
 econChart = econChart + ggtitle("Top 10 US Storm Events causing Economic Consequences") + theme(plot.title = element_text(hjust = 0.5))

 print(econChart)
```

















